---
title: "Estimating causal parameters using `geex`"
author: "Bradley Saul"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating causal parameters using geex}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
references:
- id: lunceford2004stratification
  author:
  - family: Lunceford
    given: Jared K.
  - family: Davidian
    given: Marie
  container-title: "Stratification and weighting via the propensity score in estimation of causal treatment effects: a comparative study"
  type: article-journal
  journal: Statistics in Medicine
  issued:
    year: 2004  
  volume: 23
  number: 19
  pages: 2937-2960
- id: perez-heydrich2014assessing
  author:
  - family : Perez-Heydrich
    given  : Carolina
  - family : Hudgens
    given  : Michael G.
  - family : Halloran
    given  : M. Elizabeth
  - family : Clemens
    given  : John D.
  - family : Ali
    given  : Mohammad
  - family : Emch
    given  : Michael E.
  container-title: Assessing effects of cholera vaccination in the presence of interference
  type: article-journal
  journal: Biometrics
  issued:
    year: 2014  
  volume: 70
  number: 3
  pages: 731-741
---

```{r packages, echo =TRUE}
library(geex)
library(inferference)
library(dplyr)
```

## Estimating a propensity score vs. treating propensity as known

TODO: describe what's going on here
```{r simulated_data, echo = TRUE}
n <- 1000
x <- data_frame(
  A  = rbinom(n, 1, .2),
  Y0 = rnorm(n, 0, 1),
  Y1 = rnorm(n, 2 * A, 1),
  Y = (A*Y1) + (1 - A)*Y0)
```

```{r ipw_estfun, echo = TRUE}
ipw_estFUN <- function(data){
  A <- data$A
  Y <- data$Y
  function(theta, phat){
    ipw0 <- 1/theta[1]
    ipw1 <- 1/theta[2]
    
    # Estimating functions #
    c( (1 - A) - theta[1],
      A - theta[2],
      
      # Estimating IP weight
      Y*(1 - A)*ipw0 - theta[3],
      Y*(A)*ipw1 - theta[4],  
      
      # Treating IP weight as known
      Y*A/phat - theta[5]
      )
  }
}
```

```{r ipw_estimation, echo = TRUE}
phat <- mean(x$A)
out <- m_estimate(ipw_estFUN,
           data = x,
           inner_args = list(phat = phat),
           root_control = setup_root_control(start = c(.5, .5, 0, 0, 0)))
```

```{r ipw_comparison, echo = TRUE}
## Comparing point estimates
all.equal(mean(x$Y * x$A/phat), coef(out)[4]) 
all.equal(phat, coef(out)[2]) 

## Comparing variance estimates
geex_vcov <- diag(vcov(out)) * n

# estimates match treating propensity as known
all.equal(var(x$Y * x$A/phat) * (n - 1)/n, geex_vcov[5])

# estimates match using influence function approach
y <- x$Y * x$A/phat - mean(x$Y * x$A/phat)
z <- (x$A - phat) / (phat*(1 - phat))
var(y - predict(lm(y ~ z))) - geex_vcov[4] # close
```

## IPW estimator of counterfactual mean

An example $\psi$ function written in `R`. This function computes the score functions for a GLM, plus two counterfactual means estimated by inverse probability weighting.

```{r eefun, echo=TRUE}
eefun <- function(data, model, alpha){
  X <- model.matrix(model, data = data)
  A <- model.response(model.frame(model, data = data))
  Y <- data$Y
  
  function(theta){
    p  <- length(theta)
    p1 <- length(coef(model))
    lp  <- X %*% theta[1:p1]
    rho <- plogis(lp)

    hh  <- ((rho/alpha)^A * ((1-rho)/(1-alpha))^(1 - A)) 
    IPW <- 1/(exp(sum(log(hh))))

    score_eqns <- apply(X, 2, function(x) sum((A - rho) * x))
    ce0 <- mean(Y * (A == 0)) * IPW / (1 - alpha)
    ce1 <- mean(Y * (A == 1)) * IPW / (alpha)
    
    c(score_eqns,
      ce0 - theta[p - 1],
      ce1 - theta[p])
  }
}
```

Compare to what `inferference` gets.


```{r, echo = FALSE}
if(packageVersion('inferference') < '0.5.0'){
  vaccinesim$Y <- vaccinesim$y
}
```

```{r example2, echo =TRUE}
test <- interference(
  formula = Y | A ~ X1 | group, 
  data   = vaccinesim,
  model_method = 'glm',
  allocations = c(.35, .4))

mglm <- glm(A ~ X1, data = vaccinesim, family = binomial)

ce_estimates <- m_estimate(
  estFUN = eefun,
  data  = vaccinesim,
  units = 'group',
  root_control = setup_root_control(start = c(coef(mglm), .4,  .13)),
  outer_args = list(alpha = .35, model = mglm)
)

roots(ce_estimates)

# Compare parameter estimates
direct_effect(test, allocation = .35)$estimate
roots(ce_estimates)[3] - roots(ce_estimates)[4]

# conpare SE estimates
L <- c(0, 0, 1, -1)
Sigma <- vcov(ce_estimates)
sqrt(t(L) %*% Sigma %*% L)  # from GEEX
direct_effect(test, allocation = .35)$std.error # from inferference
```


I would expect them to be somewhat different, since `inferference` uses a slightly different variance estimator defined in the [web appendix of Perez-Heydrich et al (2014)](http://onlinelibrary.wiley.com/store/10.1111/biom.12184/asset/supinfo/biom12184-sm-0001-SuppData.pdf?v=1&s=f54ecbc59e7600cd32e8f5c110d395f0a34b3fb8).

## Doubly-Robust Estimator

Estimators of causal effects often have the form: 

\begin{equation}
\label{eq:causal}
\sum_{i = 1}^m \psi(O_i, \theta) = \sum_{i = 1}^m \begin{pmatrix} \psi(O_i, \nu) \\ \psi(O_i, \beta) \end{pmatrix} = 0,
\end{equation}

\noindent where $\nu$ are parameters in nuisance model(s), such as a propensity score model, and $\beta$ are the target causal parameters. Even when $\nu$ represent parameters in common statistical models, deriving a closed form for a sandwich variance estimator for $\beta$ based on Equation~\ref{eq:causal} may involve tedious and error-prone derivative and matrix calculations [e.g.; see the appendices of @lunceford2004stratification and @perez-heydrich2014assessing]. In this example, we show how an analyst can avoid these calculations and compute the empirical sandwich variance estimator using `geex`. 

@lunceford2004stratification review several estimators of causal effects from observational data. To demonstrate a more complicated estimator involving multiple nuisance models, we implement the doubly robust estimator:

\begin{equation}
\label{eq:dbr}
\hat{\Delta}_{DR} = \sum_{i = 1}^m \frac{Z_iY_i - (Z_i - \hat{e}_i) m_1(X_i, \hat{\alpha}_1)}{\hat{e}_i} - \frac{(1 - Z_i)Y_i - (Z_i - \hat{e}_i) m_0(X_i, \hat{\alpha}_0)}{1 - \hat{e}_i}.
\end{equation}

This estimator targets the average causal effect, $\Delta = \E[Y(1) - Y(0)]$, where $Y(z)$ is the potential outcome for an experimental unit had it been exposed to the level $z$ of the binary exposure variable $Z$. The estimated propsensity score, $\hat{e}_i$, is the estimated probability that unit $i$ received $z = 1$ and $m_z(X_i, \hat{\alpha}_z)$ are regression models for the outcome with baseline covariates $X_i$ and estimated paramaters $\hat{\alpha}_z$. This estimator has the property that if either the propensity score models or the outcome models are correctly specified, then the solution to Equation~\ref{eq:dbr} will be a consistent and asymptotically Normal estimator of $\Delta$.

This estimator and its estimating equations can be translated into an `estFUN` as:

```{r dr_estfun, echo = TRUE}
dr_estFUN <- function(data, models){
  
  Z <- data$Z
  Y <- data$Y
  
  Xe <- grab_design_matrix(
    data, rhs_formula = grab_fixed_formula(models$e))
  Xm0 <- grab_design_matrix(
    data, rhs_formula = grab_fixed_formula(models$m0))
  Xm1 <- grab_design_matrix(
    data, rhs_formula = grab_fixed_formula(models$m1))
  
  e_pos  <- 1:ncol(Xe)
  m0_pos <- (max(e_pos) + 1):(max(e_pos) + ncol(Xm0))
  m1_pos <- (max(m0_pos) + 1):(max(m0_pos) + ncol(Xm1))
  
  e_scores  <- grab_psiFUN(models$e, data)
  m0_scores <- grab_psiFUN(models$m0, data)
  m1_scores <- grab_psiFUN(models$m1, data)
  
  function(theta){
    e  <- plogis(Xe %*% theta[e_pos]) 
    m0 <- Xm0 %*% theta[m0_pos] 
    m1 <- Xm1 %*% theta[m1_pos] 
    rd_hat <- (Z*Y - (Z - e) * m1)/e - ((1 - Z) * Y - (Z - e) * m0)/(1 - e)
    c(e_scores(theta[e_pos]),
      m0_scores(theta[m0_pos]) * (Z == 0),
      m1_scores(theta[m1_pos]) * (Z == 1),
      rd_hat - theta[length(theta)])  
  }
}
```

\noindent This `estFUN` presumes that the user will pass a list containing fitted model objects for the three nuisance models: the propensity score model and one regression model for each treatment group. The functions `grab_design_matrix` and `grab_fixed_formula` are `geex` utilities for extracting relevant pieces of a model object. The function `grab_psiFUN` converts a fitted model object to an estimating function; for example, for a `glm` object, `grab_psiFUN` uses the \code{data} to create a `function` of `theta` corresponding to the generalized linear model score function. The `m_estimate` function can be wrapped in another function, wherein nuisance models are fit and passed to `m_estimate`. 

```{r estimate_dr}
estimate_dr <- function(data, propensity_formula, outcome_formula){
  e_model  <- glm(propensity_formula, data = data, family = binomial)
  m0_model <- glm(outcome_formula, subset = (Z == 0), data = data)
  m1_model <- glm(outcome_formula, subset = (Z == 1), data = data)
  models <- list(e = e_model, m0 = m0_model, m1 = m1_model)
  nparms <- sum(unlist(lapply(models, function(x) length(coef(x))))) + 1
  
  m_estimate(
    estFUN = dr_estFUN,
    data   = data,
    root_control = setup_root_control(start = rep(0, nparms)),
    outer_args = list(models = models))
}
```

The following code provides a function to replicate the simulation settings of @lunceford2004stratification.

```{r lunceford_simulation, echo = TRUE}
library(mvtnorm)
tau_0 <- c(-1, -1, 1, 1)
tau_1 <- tau_0 * -1
Sigma_X3 <- matrix(
   c(1, 0.5, -0.5, -0.5,
     0.5, 1, -0.5, -0.5,
     -0.5, -0.5, 1, 0.5,
     -0.5, -0.5, 0.5, 1), ncol = 4, byrow = TRUE)

gen_data <- function(n, beta, nu, xi){
  X3 <- rbinom(n, 1, prob = 0.2)
  V3 <- rbinom(n, 1, prob = (0.75 * X3 + (0.25 * (1 - X3))))
  hold <- rmvnorm(n,  mean = rep(0, 4), Sigma_X3)
  colnames(hold) <- c("X1", "V1", "X2", "V2")
  hold <- cbind(hold, X3, V3)
  hold <- apply(hold, 1, function(x){
    x[1:4] <- x[1:4] + tau_1^(x[5])*tau_0^(1 - x[5])
    x})
  hold <- t(hold)[, c("X1", "X2", "X3", "V1", "V2", "V3")]
  X <- cbind(Int = 1, hold)
  Z <- rbinom(n, 1, prob = plogis(X[, 1:4] %*% beta))
  X <- cbind(X[, 1:4], Z, X[, 5:7])
  data.frame(
    Y = X %*% c(nu, xi) + rnorm(n), 
    X[ , -1])
}

```

To show that `estimate_dr` correctly computes $\hat{\Delta}_{DR}$, the results from `geex` can be compared to computing $\hat{\Delta}_{DR}$ "by hand" for a simulated dataset.

```{r dr_estimation}
dt <- gen_data(1000, c(0, 0.6, -0.6, 0.6), c(0, -1, 1, -1, 2), c(-1, 1, 1))
geex_results <- estimate_dr(dt, Z ~ X1 + X2 + X3, Y ~ X1 + X2 + X3)
```

```{r dr_byhand}
e  <- predict(glm(Z ~ X1 + X2 + X3, data = dt, family = "binomial"),
              type = "response")
m0 <- predict(glm(Y ~ X1 + X2 + X3, data = dt, subset = Z==0), newdata = dt)
m1 <- predict(glm(Y ~ X1 + X2 + X3, data = dt, subset = Z==1), newdata = dt)
del_hat <- with(dt, mean( (Z * Y - (Z - e) * m1)/e)) - 
  with(dt, mean(((1 - Z) * Y - (Z - e) * m0)/(1 - e)))
```
