---
title: "Estimating causal parameters using `geex`"
author: "Bradley Saul"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating causal parameters using geex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## IPW estimator of counterfactual mean

An example $\psi$ function written in `R`. This function computes the score functions for a GLM, plus two counterfactual means estimated by inverse probability weighting.

```{r eefun, echo=TRUE}
eefun <- function(data, model, alpha){
  X <- model.matrix(model, data = data)
  A <- model.response(model.frame(model, data = data))
  Y <- data$Y
  
  function(theta){
    p  <- length(theta)
    p1 <- length(coef(model))
    lp  <- X %*% theta[1:p1]
    rho <- plogis(lp)

    hh  <- ((rho/alpha)^A * ((1-rho)/(1-alpha))^(1 - A)) 
    IPW <- 1/(exp(sum(log(hh))))

    score_eqns <- apply(X, 2, function(x) sum((A - rho) * x))
    ce0 <- mean(Y * (A == 0)) * IPW / (1 - alpha)
    ce1 <- mean(Y * (A == 1)) * IPW / (alpha)
    
    c(score_eqns,
      ce0 - theta[p - 1],
      ce1 - theta[p])
  }
}
```

Compare to what `inferference` gets.

```{r packages, echo =TRUE}
library(geex)
library(inferference)
```

```{r, echo = FALSE}
if(packageVersion('inferference') < '0.5.0'){
  vaccinesim$Y <- vaccinesim$y
}
```

```{r example2, echo =TRUE}
test <- interference(
  formula = Y | A ~ X1 | group, 
  data   = vaccinesim,
  model_method = 'glm',
  allocations = c(.35, .4))

mglm <- glm(A ~ X1, data = vaccinesim, family = binomial)

ce_estimates <- m_estimate(
  estFUN = eefun,
  data  = vaccinesim,
  units = 'group',
  root_control = setup_root_control(start = c(coef(mglm), .4,  .13)),
  outer_args = list(alpha = .35, model = mglm)
)

roots(ce_estimates)

# Compare parameter estimates
direct_effect(test, allocation = .35)$estimate
roots(ce_estimates)[3] - roots(ce_estimates)[4]

# conpare SE estimates
L <- c(0, 0, 1, -1)
Sigma <- vcov(ce_estimates)
sqrt(t(L) %*% Sigma %*% L)  # from GEEX
direct_effect(test, allocation = .35)$std.error # from inferference
```


I would expect them to be somewhat different, since `inferference` uses a slightly different variance estimator defined in the [web appendix of Perez-Heydrich et al (2014)](http://onlinelibrary.wiley.com/store/10.1111/biom.12184/asset/supinfo/biom12184-sm-0001-SuppData.pdf?v=1&s=f54ecbc59e7600cd32e8f5c110d395f0a34b3fb8).
