---
output: pdf_document
---

```{r SB1_setup, echo=FALSE}
n  <- 100
mu <- 5
sigma <- 2
dt <- data.frame(Y = rnorm(n, mean = mu, sd = sigma), id = 1:n)
```

Example 1 illustrates calculation of sample mean and variance using estimating equations. I generate a data set with `r n` observations drawn from a Normal(`r mu`, `r sigma`) distribution. Table \ref{ex1} translates the estimating equations into the `R` function needed for `geex`:


\begin{table}[H]
\centering
\label{ex1}
\caption{Translating math to code}
\begin{tabular}{cc}
$\psi(Y_i, \theta) = 
\begin{pmatrix}
Y_i - \theta_1 \\
(Y_i - \theta_1)^2 - \theta_2
\end{pmatrix}$ &
\begin{minipage}{3in}
\begin{verbatim}
SB1_eefun <- function(data){
  function(theta){
    with(data,
      c(Y - theta[1],
       (Y - theta[1])^2 - theta[2] )
    )
  }
}
\end{verbatim}
\end{minipage}
\end{tabular}
\end{table}

```{r SB1_eefun, echo=FALSE, results='hide'}
SB1_eefun <- function(data){
  function(theta){
    with(data,
      c(Y - theta[1],
       (Y - theta[1])^2 - theta[2] )
    )
  }
}
```

With the `eeFUN` function prepared, it is passed to `estimate_equations` along with the data, a character string naming the variable that identifies groups within the dataset, and starting values for the root finder.

```{r SB1_run, echo=TRUE}
estimates <- estimate_equations(eeFUN = SB1_eefun, 
                                data  = dt,
                                units = 'id', 
                                roots = c(1,1))
```

```{r SB1_clsform, echo=FALSE}
## Compare to closed form ##

A <- diag(1, nrow = 2)

B <- with(dt, {
  Ybar <- mean(Y)
  B11 <- mean( (Y - Ybar)^2 )
  B12 <- mean( (Y - Ybar) * ((Y - Ybar)^2 - B11) )
  B22 <- mean( ((Y - Ybar)^2 - B11)^2 )
  matrix(
    c(B11, B12,
      B12, B22), nrow = 2
  )
})

## closed form roots
# note that var() divides by n - 1, not n
theta_cls <- summarize(dt, p1 = mean(Y), p2 = var(Y) * (n() - 1)/ n() )

# closed form
Sigma_cls <- (solve(A) %*% B %*% t(solve(A))) / n
```

```{r SB1_results, echo = FALSE, results = 'asis'}
results <- list(geex = estimates, cls = list(parameters = theta_cls, vcov = Sigma_cls))

print_results(results, 'ex1', 'Comparing estimates from closed form versus geex')
```
