---
output: pdf_document
---


```{r SB3_setup, echo=FALSE}
library(geex)
library(dplyr)
n  <- 100
mu <- 5
sigma <- 4
set.seed(100) # running into issue where sqrt(theta2) and log(theta2) return NaN for some seeds
dt <- data.frame(Y  = rnorm(n, mean = mu, sd = sigma), 
                 id = 1:n)
```

Example 3 illustrates calculation of a ratio estimator. I generate a data set with `r n` observations where $Y \sim N$(`r mu`, `r sigma`). Table \ref{ex3} translates the estimating equations into the `R` function needed for `geex`:


\begin{table}[H]
\centering
\label{ex3}
\caption{Translating math to code}
\begin{tabular}{cc}
$\psi(Y_i, \theta) = 
\begin{pmatrix}
Y_i - \theta_1 \\
X_i - \theta_2 \\
\sqrt{\theta_2} - \theta_3 \\
log(\theta_2) - \theta_4
\end{pmatrix}$ &
\begin{minipage}{3in}
\begin{verbatim}
SB3_eefun <- function(data){
  function(theta){
    with(data,
      c(Y - theta[1],
       (Y - theta[1])^2 - theta[2],
       sqrt(theta[2]) - theta[3],
       log(theta[2]) - theta[4])
    )
  }
}
\end{verbatim}
\end{minipage}
\end{tabular}
\end{table}

```{r SB3_eefun, echo = FALSE}
SB3_eefun <- function(data){
  function(theta){
    with(data,
      c(Y - theta[1],
       (Y - theta[1])^2 - theta[2],
       sqrt(theta[2]) - theta[3],
       log(theta[2]) - theta[4])
    )
  }
}
```

```{r SB3_run, echo = TRUE}
estimates <- estimate_equations(eeFUN= SB3_eefun, 
                                data  = dt, units = 'id', 
                                roots = c(1, 1, 1, 1))
```

```{r SB3_clsform, echo = FALSE}
## closed form roots
theta_cls <- summarize(dt, p1 = mean(Y), p2 = sum((Y - p1)^2)/n(), p3 = sqrt(p2), p4 = log(p2))

## Compare to closed form ##
theta2 <- theta_cls$p2
mu3 <- moments::moment(dt$Y, order = 3, central = TRUE)
mu4 <- moments::moment(dt$Y, order = 4, central = TRUE)
# A <- matrix(c(1, 0, 0, 0,
#               0, 1, 0, 0,
#               0, -1/(2 * sqrt(theta2)), 1, 0,
#               0, -1/theta2, 0, 1), 
#             byrow = TRUE, nrow = 4)
# B <- matrix(c(1/theta2, mu3/(2 * theta2^3), 0, 0,
#               mu3/(2 * theta2^3), (mu4 - theta2^2)/(4 * theta2^4), 0, 0,
#               0, 0, 0, 0,
#               0, 0, 0, 0),
#             byrow = TRUE, nrow = 4)

## closed form covariance
Sigma_cls <- matrix(
  c(theta2, mu3, mu3/(2*sqrt(theta2)), mu3/theta2,
    mu3, mu4 - theta2^2, (mu4 - theta2^2)/(2*sqrt(theta2)), (mu4 - theta2^2)/theta2,
    mu3/(2 * sqrt(theta2)), (mu4 - theta2^2)/(2*sqrt(theta2)), (mu4 - theta2^2)/(4*theta2), (mu4 - theta2^2)/(2*theta2^(3/2)),
    mu3/theta2, (mu4 - theta2^2)/theta2, (mu4 - theta2^2)/(2*theta2^(3/2)), (mu4/theta2^2) - 1) ,
  nrow = 4, byrow = TRUE) / n
## closed form covariance
# Sigma_cls <- (solve(A) %*% B %*% t(solve(A))) / n
```


```{r SB3_results, echo = FALSE, results = 'asis'}
results <- list(geex = estimates, cls = list(parameters = theta_cls, vcov = Sigma_cls))
results
# print_results(results, 'ex3_results', 'Example 3')
```
