---
output: pdf_document
---

```{r SB9_setup, echo=FALSE}
library(geex)
library(dplyr)
library(sandwich)
n <- 100
beta <- c(0.5, 2, .1)
dt <- data_frame(X1 = rep(0:1, each = n/2), 
                 X2 = rep(0:1, times = n/2),
                 Y  = rbinom(n, 1, prob = as.numeric(plogis(cbind(1, X1, X2) %*% beta))),
                 id = 1:n)
```

Example 9 illustrates estimation of a generalized linear model. I generate a data set with `r n` observations where half of the observation have $X_{1i} = 1$ and the others have $X_{1i} = 0$. $Y_i \sim Bern[\mbox{logit}^{-1}($ `r beta[1]` $+$ `r beta[2]` $X_{1i} +$ `r beta[3]` $X_{2i})]$. Table \ref{ex9} translates the estimating equations for logistic regression into the `R` function needed for `geex`:

\begin{table}[H]
\centering
\label{ex9}
\caption{Translating math to code}
\begin{tabular}{cc}
$\psi_k(Y_i, \theta) = 
\begin{pmatrix}
D_i(\beta)\frac{Y_i - \mu_i(\beta)}{V_i(\beta) \tau}
\end{pmatrix}$ &
\begin{minipage}{3in}
\begin{verbatim}
SB9_eefun <- function(data){
  Yi <- data$Y
  xi <- model.matrix(Y ~ X1 + X2, data = data, drop = FALSE)
  function(theta){
    lp <- xi %*% theta
    mu <- plogis(lp)
    D  <- t(xi) %*% dlogis(lp)
    V  <- mu * (1 - mu)
    D %*% solve(V) %*% (Yi - mu)
  }
}
\end{verbatim}
\end{minipage}
\end{tabular}
\end{table}

```{r SB9_eefun, echo = FALSE}
SB9_eefun <- function(data){
  Yi <- data$Y
  xi <- model.matrix(Y ~ X1 + X2, data = data, drop = FALSE)
  function(theta){
    lp <- xi %*% theta
    mu <- plogis(lp)
    D  <- t(xi) %*% dlogis(lp)
    V  <- mu * (1 - mu)
    D %*% solve(V) %*% (Yi - mu)
  }
}
```

```{r SB9_run, echo = TRUE}
estimates <- estimate_equations(eeFUN = SB9_eefun, 
                                data  = dt, units = 'id', 
                                roots = c(1, 1, 1))
```

```{r SB9_clsform, echo = TRUE}
m <- glm(Y ~ X1 + X2, data = dt, family = binomial(link = 'logit'))
theta_cls <- coef(m)
Sigma_cls <- sandwich(m)

```

```{r SB9_results, echo = FALSE, results = 'asis'}
results <- list(geex = estimates, cls = list(parameters = theta_cls, vcov = Sigma_cls))
results
# print_results(results, 'Example 9', 'Example 9')
```
