---
output: pdf_document
---

```{r SB10_setup, echo=FALSE}
shaq <- data_frame(game = 1:23,
                 ft_made = c(4, 5, 5, 5, 2, 7, 6, 9, 4, 1, 13, 5, 6, 9, 7, 3, 8, 1, 18, 3, 10, 1, 3),
                 ft_attp = c(5, 11, 14, 12, 7, 10, 14, 15, 12, 4, 27, 17, 12, 9, 12, 10, 12, 6, 39, 13, 17, 6, 12))

```

Example 10 illustrates testing equality of success probablities. Table \ref{ex10} translates the estimating equations into the `R` function needed for `geex`:

\begin{table}[H]
\centering
\label{ex10}
\caption{Translating math to code}
\begin{tabular}{cc}
$\psi_k(Y_i, n_i, \theta) = 
\begin{pmatrix}
\frac{(Y_i - n_i \theta_2)^2}{n_i \theta_2( 1 - \theta_2 )} - \theta_1 \\
Y_i - n_i \theta_2
\end{pmatrix}$ &
\begin{minipage}{3in}
\begin{verbatim}
SB10_eefun <- function(data){
  Y <- data$ft_made
  n <- data$ft_attp
  function(theta){
    p <- theta[2]
    c(((Y - (n * p))^2)/(n * p * (1 - p))  - theta[1], 
      Y - n * p)
  }
}
\end{verbatim}
\end{minipage}
\end{tabular}
\end{table}

```{r SB10_eefun, echo = FALSE}
SB10_eefun <- function(data){
  Y <- data$ft_made
  n <- data$ft_attp
  function(theta){
    p <- theta[2]
    c(((Y - (n * p))^2)/(n * p * (1 - p))  - theta[1], 
      Y - n * p)
  }
}
```

```{r SB10_run, echo = TRUE}
estimates <- estimate_equations(eeFUN = SB10_eefun, 
                                data  = shaq, units = 'game', 
                                roots = c(.5, .5))
```

```{r SB10_clsform, echo = TRUE}
V11 <- function(p) {
  k <- length(nrow(shaq))
  sumn <- sum(shaq$ft_attp)
  sumn_inv <- sum(1/shaq$ft_attp)
  term2_n <- 1 - (6 * p) + (6 * p^2)
  term2_d <- p * (1 - p) 
  term2 <- term2_n/term2_d
  print(term2)
  term3 <- ((1 - 2 * p)^2)/( (sumn/k) * p * (1 - p))
  print(term3)
  2 + (term2 * (1/k) * sumn_inv)  - term3
}

### ???? I keep getting a negative value for V11

p_tilde <- sum(shaq$ft_made)/sum(shaq$ft_attp)
V <- V11(.45)
V
pnorm(estimates$parameters[1], mean = 1, sd = sqrt(V))

```

```{r SB10_results, echo = FALSE, results = 'asis', eval = FALSE}
results <- list(geex = estimates, cls = list(parameters = theta_cls, vcov = Sigma_cls))
print_results(results, 'Example 10', 'Example 10')
```
