---
output: pdf_document
---


```{r SB_setup, echo=FALSE}
n  <- 100
muY <- 5
sigmaY <- 2
muX <- 2
sigmaX <- 0.2
dt <- data.frame(Y  = rnorm(n, mean = muY, sd = sigmaY), 
                 X  = rnorm(n, mean = muX, sd = sigmaX),
                 id = 1:n)
```

Example 2 illustrates calculation of a ratio estimator. I generate a data set with `r n` observations where $Y \sim N$(`r muY`, `r sigmaY`) and $X \sim N$(`r muX`, `r sigmaX`). Table \ref{ex2} translates the estimating equations into the `R` function needed for `geex`:


\begin{table}[H]
\centering
\label{ex2}
\caption{Translating math to code}
\begin{tabular}{cc}
$\psi(Y_i, \theta) = 
\begin{pmatrix}
Y_i - \theta_1 \\
X_i - \theta_2 \\
\theta_1 - \theta_3\theta_2
\end{pmatrix}$ &
\begin{minipage}{3in}
\begin{verbatim}
SB2_eefun <- function(data){
  function(theta){
    with(data,
      c(Y - theta[1],
        X - theta[2],
        theta[1] - (theta[3] * theta[2]) )
    )
  }
}
\end{verbatim}
\end{minipage}
\end{tabular}
\end{table}

```{r SB2_eefun, echo = FALSE}
SB2_eefun <- function(data){
  function(theta){
    with(data,
      c(Y - theta[1],
        X - theta[2],
        theta[1] - (theta[3] * theta[2]) )
    )
  }
}
```

```{r SB2_run, echo = TRUE}
estimates <- estimate_equations(eeFUN = SB2_eefun, 
                                data  = dt, units = 'id', 
                                roots = c(1, 1, 1))
```

```{r SB2_clsform, echo = FALSE}
## Compare to closed form ##

A <- with(dt, {
 matrix(
  c(1 , 0, 0,
    0 , 1, 0,
    -1, mean(Y)/mean(X), mean(X)),
  byrow = TRUE, nrow = 3)
})

B <- with(dt, {
  matrix(
    c(var(Y)   , cov(Y, X), 0,
      cov(Y, X), var(X)   , 0,
      0, 0, 0),
    byrow = TRUE, nrow = 3)
})

## closed form roots
theta_cls <- summarize(dt, p1 = mean(Y), p2 = mean(X), p3 = p1/p2)

## closed form covariance
Sigma_cls <- (solve(A) %*% B %*% t(solve(A))) / n
```

```{r SB2_results, echo = FALSE, results = 'asis'}
results <- list(geex = estimates, cls = list(parameters = theta_cls, vcov = Sigma_cls))
print_results(results, 'test', 'test')
```
