---
output: pdf_document
---


```{r SB4_setup, echo=FALSE}
n  <- 100

# Oracle parms
alpha <- 2
beta  <- 3
gamma <- 2
delta <- 1.5
e1 <- e2 <- e3 <- rnorm(n)
sigma_e <- 1
sigma_U <- .25
sigma_tau <- 1
### Random variables

X <- rgamma(n, shape = 5)
X <- rnorm(n, sd = 1)

dt <- data.frame(Y  = alpha + (beta * X) + (sigma_e * e1), 
                 W  = X + (sigma_U * e2),
                 T_  = gamma + (delta * X) + (sigma_tau * e3),
                 id = 1:n)
```

Example 4 illustrates calculation of an instumental variable estimator. I generate a data set with `r n` observations where INPUT data generation. Table \ref{ex4} translates the estimating equations into the `R` function needed for `geex`:


\begin{table}[ht]
\centering
\label{ex4}
\caption{Translating math to code}
\begin{tabular}{cc}
$\psi(Y_i, T_i, W_i \theta) = 
\begin{pmatrix}
\theta_1 - T_i\\
\theta_2 - W_i \\
(Y_i - \theta_3 W_i)(\theta_2 - W_i) \\
(Y_i - \theta_4 W_i)(\theta_1 - T_i) \\
\end{pmatrix}$ &
\begin{minipage}{3in}
\begin{verbatim}
SB_eefun <- function(data){
  function(theta){
    with(data,
      c(theta[1] - T_,
        theta[2] - W,
        (Y - (theta[3] * W)) * (theta[2] - W),
        (Y - (theta[4] * W)) * (theta[1] - T_))
    )
  }
}
\end{verbatim}
\end{minipage}
\end{tabular}
\end{table}

```{r SB4_eefun, echo = FALSE}
SB4_eefun <- function(data){
  function(theta){
    with(data,
      c(theta[1] - T_,
        theta[2] - W,
        (Y - (theta[3] * W)) * (theta[2] - W),
        (Y - (theta[4] * W)) * (theta[1] - T_))
    )
  }
}
```

```{r SB4_run, echo = TRUE}
estimates <- estimate_equations(eeFUN = SB4_eefun, 
                                data  = dt, units = 'id', 
                                roots = c(1, 1, 1, 1))
```

```{r SB4_clsform, echo = TRUE}
YW_model <- lm(Y ~ W, data = dt)
YT_model <- lm(Y ~ T_, data = dt)
WT_model <- lm(W ~ T_, data = dt)
## closed form roots
theta_cls <- c(theta1 = mean(dt$T_),
  theta2 = mean(dt$W),
  theta3 = coef(YW_model)[2],
  theta4 = coef(YT_model)[2]/coef(WT_model)[2])

## closed form covariance
# Not sure how compute SB's closed form since it depends on X, which is
# supposed to be unobserved.
Sigma_cls <- matrix(NA, nrow = 2, ncol = 2)
```

```{r SB4_results, echo = FALSE, results = 'asis'}
# primary interest lies in the lower 2 x 2 submatrix of the asymptotic variance matrix
estimates$vcov <- estimates$vcov[3:4, 3:4]
results <- list(geex = estimates, cls = list(parameters = theta_cls, vcov = Sigma_cls))
print_results(results, 'Example 4', 'Example 4')
```
