---
output: pdf_document
---


```{r SB5_setup, echo=FALSE}
n <- 100
theta0 <- 0
theta_tru <- 2
sigma <- 1
dt <- data.frame(X = rnorm(n, mean = 2, sd = sigma),
                 id = 1:n)
```

Example 5 illustrates calculation of an instumental variable estimator. I generate a data set with `r n` observations where $X \sim N$(`r theta_tru`, `r sigma`). Let $\theta_0 =$ `r theta0`. Table \ref{ex5} translates the estimating equations for the Hodges-Lehmann location estimation and the sample mean into the `R` function needed for `geex`:

\begin{table}[!ht]
\centering
\label{ex5}
\caption{Translating math to code}
\begin{tabular}{cc}
$\psi(Y_i, \theta) = 
\begin{pmatrix}
IC_{\hat{\theta}_{HL}}(X; \theta_0) - (\theta_1 - \theta_0) \\
X_i - \theta_2 \\
\end{pmatrix}$ &
\begin{minipage}{3in}
\begin{verbatim}
SB5_eefun <- function(data, theta0 = 0){
  Xi <- data$X
  IC_HL <- (1/IC_denom) * (F0(Xi, theta0) - 0.5)
  function(theta){
     c(IC_HL - (theta[1] - theta0), 
       Xi - theta[2]) 
  }
}
\end{verbatim}
\end{minipage}
\end{tabular}
\end{table}

```{r SB5_eefun, echo = TRUE}
F0 <- function(y, theta0, distrFUN = pnorm){
  distrFUN(y - theta0, mean = 0)
}

f0 <- function(y, densFUN){
  densFUN(y, mean = 0)
}

integrand <- function(y, densFUN = dnorm){
  f0(y, densFUN = densFUN)^2
}

IC_denom <- integrate(integrand, lower = -Inf, upper = Inf)$value

SB5_eefun <- function(data, theta0 = 0){
  Xi <- data$X
  IC_HL <- (1/IC_denom) * (F0(Xi, theta0) - 0.5)
  function(theta){
     c(IC_HL - (theta[1] - theta0),
       Xi - theta[2]) 
  }
}
```

```{r SB5_run, echo = TRUE}
estimates <- estimate_equations(eeFUN = SB5_eefun, 
                                data  = dt, units = 'id', 
                                roots = c(1, 1))
```

```{r SB5_clsform, echo = TRUE}
X <- dt$X
pair_means <- numeric(length(X) - 1)
for(i in 1:(length(X) - 1)){
 pair_means[i] <-  (X[i] + X[i + 1])/2
}

theta_cls <- c(median(pair_means), mean(X))

## closed form covariance
# Not sure how compute SB's closed form since it depends on X, which is
# supposed to be unobserved.
Sigma_cls <- matrix(c(1/(12 * IC_denom^2) / n, NA, NA, NA), 
                    nrow = 2, ncol = 2, byrow = TRUE)
```

```{r SB5_results, echo = FALSE, results = 'asis'}
results <- list(geex = estimates, cls = list(parameters = theta_cls, vcov = Sigma_cls))
print_results(results, 'Example 5', 'Example 5')
```
