---
output: pdf_document
---


```{r SB6_setup, echo=FALSE}
n <- 100
theta_tru <- 2
sigma <- 1
dt <- data.frame(Y = rnorm(n, mean = 2, sd = sigma),
                 id = 1:n)
```

Example 6 illustrates calculation of an instumental variable estimator. I generate a data set with `r n` observations where $Y \sim N$(`r theta_tru`, `r sigma`). Table \ref{ex6} translates the estimating equations for the Huber estimator for the center of symmetric distributions into the `R` function needed for `geex`:

\begin{table}[!ht]
\centering
\label{ex6}
\caption{Translating math to code}
\begin{tabular}{cc}
$\psi_k(Y_i, \theta) = 
\begin{pmatrix}
(Y_i - \theta) * I(|(Y_i - \theta)| \leq k) + k * sgn(Y_i - \theta)\\
\end{pmatrix}$ &
\begin{minipage}{3in}
\begin{verbatim}
SB6_eefun <- function(data, k = 1.345){
  function(theta){
    x <- data$Y - theta[1]
    if(abs(x) <= k) x else sign(x) * k
  }
}
\end{verbatim}
\end{minipage}
\end{tabular}
\end{table}

```{r SB6_eefun, echo = TRUE}
SB6_eefun <- function(data, k = 1.5){
  function(theta){
    x <- data$Y - theta[1]
    if(abs(x) <= k) x else sign(x) * k
  }
}
```

```{r SB6_run, echo = TRUE}
estimates <- estimate_equations(eeFUN = SB6_eefun, 
                                data  = dt, units = 'id', 
                                roots = 1)
```

```{r SB6_clsform, echo = TRUE}
library(MASS)
theta_cls <- MASS::huber(dt$Y)$mu

## closed form covariance: TODO
Sigma_cls <- matrix(NA)
```

```{r SB6_results, echo = FALSE, results = 'asis'}
results <- list(geex = estimates, cls = list(parameters = theta_cls, vcov = Sigma_cls))
print_results(results, 'Example 6', 'Example 6')
```
