---
output: pdf_document
---

```{r SB7_setup, echo=FALSE}
n <- 100
theta_tru <- 2
sigma <- 1
dt <- data.frame(Y = rnorm(n, mean = 2, sd = sigma),
                 id = 1:n)
```

Example 7 illustrates calculation of sample quantiles using M-estimation. I generate a data set with `r n` observations where $Y \sim N$(`r theta_tru`, `r sigma`). Table \ref{ex7} translates the estimating equations for two sample quantiles (median and 65th percentile) into the `R` function needed for `geex`:

\begin{table}[!ht]
\centering
\label{ex7}
\caption{Translating math to code}
\begin{tabular}{cc}
$\psi_k(Y_i, \theta) = 
\begin{pmatrix}
0.5 - I(Y_i \leq \theta_1) \\
0.65 - I(Y_i \leq \theta_2) 
\end{pmatrix}$ &
\begin{minipage}{3in}
\begin{verbatim}
SB7_eefun <- function(data){
  function(theta){
    with(data,
      c(0.5  - (Y <= theta[1]),
        0.65 - (Y <= theta[2]))
    )
  }
}
\end{verbatim}
\end{minipage}
\end{tabular}
\end{table}

```{r SB7_eefun, echo = TRUE}
SB7_eefun <- function(data){
  function(theta){
    with(data,
      c(0.5  - (Y <= theta[1]),
        0.65 - (Y <= theta[2]))
    )
  }
}
```

```{r SB7_run, echo = TRUE, eval=FALSE}
estimates <- estimate_equations(eeFUN = SB7_eefun, 
                                data  = dt, units = 'id', 
                                roots = c(.5, .65))
```

```{r SB7_clsform, echo = TRUE}
theta_cls <- c(quantile(dt$Y, 0.5), quantile(dt$Y, 0.65))

```

```{r SB7_results, echo = FALSE, results = 'asis', eval=FALSE}
results <- list(geex = estimates, cls = list(parameters = theta_cls, vcov = Sigma_cls))
print_results(results, 'Example 7', 'Example 7')
```
