---
output: pdf_document
---

```{r SB8_setup, echo=FALSE}
n <- 50
beta <- c(0.5, 2)
dt <- data_frame(X  = rep(0:1, each = n/2),
                 e  = rnorm(n),
                 Y  = as.numeric(cbind(1, X) %*% beta) + e,
                 id = 1:n)
```

Example 8 illustrates robust regression. I generate a data set with `r n` observations where half of the observation have $X_i = 1$ and the others have $X_i = 0$. $Y = $ `r beta[1]` $+$ `r beta[2]` $X_i + \epsilon_i$ and $\epsilon_i \sim N(0, 1)$. Table \ref{ex8} translates the estimating equations for two sample quantiles (median and 65th percentile) into the `R` function needed for `geex`:

\begin{table}[H]
\centering
\label{ex8}
\caption{Translating math to code}
\begin{tabular}{cc}
$\psi_k(Y_i, \theta) = 
\begin{pmatrix}
\psi_k(Y_i - \mathbf{x}_i^T \beta) \mathbf{x}_i
\end{pmatrix}$ &
\begin{minipage}{3in}
\begin{verbatim}
SB8_eefun <- function(data){
  Yi <- data$Y
  xi <- model.matrix(Y ~ X, data = data)
  function(theta){
    r <- Yi - xi %*% theta
    c(psi_k(r) %*% xi)
  }
}
\end{verbatim}
\end{minipage}
\end{tabular}
\end{table}

```{r SB8_eefun, echo = FALSE}
psi_k <- function(x, k = 1.345){
  if(abs(x) <= k) x else sign(x) * k
}

SB8_eefun <- function(data){
  Yi <- data$Y
  xi <- model.matrix(Y ~ X, data = data)
  function(theta){
    r <- Yi - xi %*% theta
    c(psi_k(r) %*% xi)
  }
}
```

```{r SB8_run, echo = TRUE}
estimates <- estimate_equations(eeFUN = SB8_eefun, 
                                data  = dt, units = 'id', 
                                roots = c(1, 1))
```

```{r SB8_clsform, echo = TRUE}
m <- MASS::rlm(Y ~ X, data = dt, method = 'M')
theta_cls <- coef(m)
Sigma_cls <- vcov(m)

```

```{r SB8_results, echo = FALSE, results = 'asis'}
results <- list(geex = estimates, cls = list(parameters = theta_cls, vcov = Sigma_cls))
print_results(results, 'Example 8', 'Example 8')
```
