---
title: "Tips for making geex faster"
output: html_document
---

*Under development*

## Do everything not involving theta outside of the psiFUN

The following example compares two `estFUN`s. The first performs manipulations of `data` outside of the "inner" `psiFUN` function. The second performs the manipulations within the returned `psiFUN`.
```{r}

glm_eefun1 <- function(data, model){
  f <- grab_psiFUN(model, data)
  function(theta){
    f(theta)
  }
}

glm_eefun2 <- function(data, model){
  function(theta){
    f <- grab_psiFUN(model, data)
    f(theta)
  }
}
```

Now, compare the speed:

```{r setup}
library(geex)
library(geepack, quietly = TRUE)
data("ohio")

test_binomial <- glm(resp ~ age, data = ohio,
                     family = binomial(link = "logit"))

system.time({test <- m_estimate(
  glm_eefun1,
  data = ohio,
  units = 'id',
  root_control  = setup_root_control(start = coef(test_binomial)),
  outer_args = list(model = test_binomial))})

system.time(update(test, estFUN = glm_eefun2))
```

## Use `method = "simple"` for computing numerical derivatives

The default for the function `numDeriv::jacobian` uses the "Richardson" method for computing numerical derivatives, which requires multiple evaluations of each `psiFUN`. The "simple" method is much faster, though less accurate. See the `numDeriv::jacobian` help page for further options.

```{r}
# The default uses method = 'Richardson`
system.time({test <- m_estimate(
  estFUN = glm_eefun1, data = ohio, units = "id",
  roots = coef(test_binomial),
  compute_roots = FALSE,
  outer_args = list(model = test_binomial))})

# Using method = "simple"
system.time(update(test, 
          deriv_control = setup_deriv_control(method = "simple")))
```

