---
title: "Causal example"
author: "B. Saul"
date: "`r Sys.Date()`"
output: pdf_document
header-includes:
- \usepackage{float}
vignette: >
  %\VignetteIndexEntry{geex intro}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## Example: comparison to `sandwich`

I'll use the `vaccinesim` dataset.


An example $\psi$ function written in `R`. This function computes the score functions for a GLM.
```{r eefun, echo=TRUE}
eefun <- function(data, model){
  X <- model.matrix(model, data = data)
  Y <- model.response(model.frame(model, data = data))
  function(theta){
    lp  <- X %*% theta
    rho <- plogis(lp)

    score_eqns <- apply(X, 2, function(x) sum((Y - rho) * x))
    score_eqns
  }
}
```

Compare sandwich variance estimators to `sandwich` treating individuals as units:
```{r example1}
library(inferference)
vaccinesim$ID <- 1:nrow(vaccinesim)
mglm    <- glm(A ~ X1, data = vaccinesim, family = binomial)
split_data  <- split(vaccinesim, vaccinesim$ID)
# The list needed for the compute_matrices
# For now, theta needs to be passed since geex does not do point estimates yet
example <- list(eeFUN = eefun, splitdt = split_data)
estimates <- estimate_equations(eeFUN = eefun,
                   data = vaccinesim,
                   units = 'ID',
                   roots = c(-.35, 0),
                   model = mglm)

# Compare point estimates
estimates$parameters # from GEEX
coef(mglm) # from the GLM function

# Compare variance estimates
estimates$vcov
sandwich::sandwich(mglm)
```

Pretty darn good! Note that the `geex` method is much slower than `sandwich` (especially using `method = 'Richardson'` for `numDeriv`), but this is because `sandwich` uses the closed form of the score equations, while `geex` compute them numerically. However, `geex`'s real utility comes when you have more complicated estimating equations. Also, the analyst has the ability to code faster $\psi$ functions by optimizing their code or using `Rccp`, for example. 

## Example: IPW estimator of counterfactual mean

An example $\psi$ function written in `R`. This function computes the score functions for a GLM, plus two counterfactual means estimated by inverse probability weighting.

```{r eefun2, echo=TRUE}
eefun2 <- function(data, model, alpha){
  X <- model.matrix(model, data = data)
  A <- model.response(model.frame(model, data = data))
  
  function(theta){
    p  <- length(theta)
    p1 <- length(coef(model))
    lp  <- X %*% theta[1:p1]
    rho <- plogis(lp)

    hh  <- ((rho/alpha)^A * ((1-rho)/(1-alpha))^(1 - A)) 
    IPW <- 1/(exp(sum(log(hh))))

    score_eqns <- apply(X, 2, function(x) sum((A - rho) * x))
    with(data, {
      ce0 <- mean(Y * (A == 0)) * IPW / (1 - alpha)
      ce1 <- mean(Y * (A == 1)) * IPW / (alpha)
      
      c(score_eqns,
        ce0 - theta[p - 1],
        ce1 - theta[p])
    })
  }
}
```

Compare to what `inferference` gets.

```{r example2, echo =TRUE}
test <- interference(Y | A ~ X1 | group, 
                     data = vaccinesim,
                     model_method = 'glm',
                     allocations = c(.35, .4))

mglm        <- glm(A ~ X1, data = vaccinesim, family = binomial)
split_data  <- split(vaccinesim, vaccinesim$group)

# The list needed for the compute_matrices
example <- list(eeFUN   = eefun2, 
                splitdt = split_data)

ce_estimates <- estimate_equations(
  eeFUN = eefun2,
  data  = vaccinesim,
  units = 'group',
  roots = c(coef(mglm), .4,  .13),
  alpha = .35,
  model = mglm
)

ce_estimates$parameters

# Compare parameter estimates
direct_effect(test, allocation = .35)$estimate
ce_estimates$parameters[3] - ce_estimates$parameters[4]

# conpare SE estimates
L <- c(0, 0, 1, -1)
Sigma <- ce_estimates$vcov
sqrt(t(L) %*% Sigma %*% L)  # from GEEX
direct_effect(test, allocation = .35)$std.error # from inferference
```


I would expect them to be somewhat different, since `inferference` computes the variance with the block diagonal trick in the Perez appendix.
